\chapter{Introduction} 

\section{Livestock farming, animal health and infectious diseases}

\subsection{Socio-economical context and the stakes}

Livestock farming is a cornerstone of agricultural systems, essential for food production and the livelihood of many communities. However, modern farming faces a dual challenge: ensuring high levels of animal health and well-being while transitioning toward sustainable, self-sufficient practices. In this section, I will first give a general context on livestock farming , ...
I will outline 
%the socio-economic impacts our work could have.


\subsubsection{Livestock Farming, animal health well-being}

First, we define the core concepts:
\begin{itemize}
    \item Livestock Farming: The organized breeding and raising of animals for food, fiber, and other products.
    \item Animal Health and Well-Being: The physical and psychological state of animals, which directly impacts productivity and the quality of the products.
    \item Sustainable/Self-Sufficient Livestock Farming: Farming systems that are designed to maintain productivity and animal health over the long term with minimal reliance on external inputs, promoting environmental, economic, and social resilience.
\end{itemize}


Next, we examine why sustainable livestock farming is critical in today’s global context:

\begin{itemize}
    \item Food Security: As the global population grows, the demand for high-quality animal products increases. Ensuring that farming practices are sustainable helps guarantee that future generations will have access to sufficient, nutritious food. This involves not only meeting quantity demands but also maintaining quality standards.
    \item Animal Welfare and public Health: The safety of the food supply is paramount. Sustainable practices reduce the risks of diseases that can be transmitted between animals and humans. Maintaining robust animal health also minimizes the use of interventions, such as antibiotics, which can lead to issues like antibiotic resistance.
    \item Environmental and Ecological Management: Livestock farming significantly contributes to environmental challenges such as greenhouse gas emissions, pollution, and deforestation. Transitioning to sustainable systems can mitigate these impacts, balancing food production with the need to preserve natural ecosystems.
    \item Resource Management: Efficient farming requires the optimal use of resources—land, water, energy, and labor. Sustainable systems are designed to reduce waste and improve efficiency. This is particularly important as both the availability of resources and the demand for skilled labor (veterinarians, farmers, technicians) face increasing pressures in many parts of the world.
\end{itemize}

By addressing these interrelated aspects, sustainable livestock farming not only tackles immediate socio-economic challenges but also contributes to long-term ecological balance and public health protection.

\textit{\textbf{Keywords:}} Livestock farming, animal health, animal-welfare, resource management, food security, ecology


\subsubsection{Infectious diseases - impact on livestock farming}

Infectious diseases are disorders caused by pathogenic microorganisms—such as bacteria, viruses, fungi, or parasites—that can spread directly or indirectly from one host to another. These diseases can manifest across different domains, affecting both humans and animals, and their transmission may occur at multiple scales—from within a single host to entire metapopulations.

\textbf{Types and Transmission Patterns:} Infectious diseases vary by type. For instance, respiratory infections (e.g., COVID-19 in humans and certain respiratory illnesses in animals) and sexually transmitted diseases (such as AIDS in humans) illustrate the diversity of these conditions. In the animal domain, examples like African swine fever and bovine viral diarrhea underscore how pathogens can devastate livestock. Some infections are zoonotic, meaning they can jump from wildlife to livestock and even to humans (e.g., Ebola, monkeypox), highlighting the interconnectedness of human, animal, and environmental health. Transmission can occur in several ways, including host-to-host contact, within-host dynamics, and across populations, which sometimes creates complex patterns that pose significant public health challenges.

The effects of infectious diseases are far-reaching:
\begin{itemize}
    \item Ecological Impact: Inappropriate or excessive use of antimicrobials in response to infections can lead to the emergence of resistant bacteria. This resistance not only renders certain treatments ineffective but also complicates recovery, thereby jeopardizing both animal and human health.
    \item Food Safety and Security: Outbreaks can cause major production losses in livestock farming by reducing both the quality and quantity of animal products. These losses threaten the broader goal of achieving food security, especially in regions where livestock is a primary source of nutrition and income
    \item Animal Health and Welfare: Infectious diseases directly compromise the health and well-being of animals, leading to suffering and significant economic losses for farmers.
    \item Resource Management: Managing infectious diseases demands frequent monitoring and diagnosis, along with substantial investments in vaccines, antibiotics, and veterinary expertise. This imposes a heavy financial and operational burden on farming systems, making sustainable practices even more challenging to implement.
\end{itemize}

By understanding these aspects, one recognizes that controlling infectious diseases is not only a matter of public health but also a critical component of creating sustainable and self-sufficient livestock farming systems.

\textit{\textbf{Keywords:}} infectious diseases, critical concern, animal welfare, animal health, antimicrobial resistance, economical loss, production loss, food safety, public health, epidemiology

% Did I miss key ideas that could help me demonstrate why infectious diseases in livestock farming are interresting topics in society currently ? 
% Also i think inn this subsection i need to give more example of infectious diseases in the livestok farming context. The other ones (for humans, or wild animals) are juste to give an idea or illustrate something. 

\subsection{Conventional control strategies of infectious diseases}
\label{Conventional control strategies} % ajout-du-label-pour-faire-reference-a-la-sous-section

\subsubsection{Diagnose, prognose and recommend control measures}

Effective management of infectious diseases in livestock relies on a systematic process that begins with accurate diagnosis, proceeds with reliable prognosis, and culminates in tailored recommendations for control measures. Each step is crucial for mitigating outbreaks and ensuring the long-term sustainability of farming systems.

\paragraph{Diagnosis.} The first step involves identifying the presence and nature of the disease. Diagnosis can be achieved through:
\begin{itemize}
    \item Clinical Appraisal: This is the most common method among farmers, where visual or tactile observations are used to detect clinical signs in animals. While this approach is quick and accessible, it is inherently subjective and may lead to inconsistent interpretations.
    \item Biological Examinations: In this approach, samples such as blood or mucus are collected for laboratory analysis (e.g., PCR testing). Although this method can provide a more precise identification of pathogens, it is often more invasive, time-consuming, and may delay decision-making during fast-moving outbreaks.
\end{itemize}

\paragraph{Prognosis.} Once the disease is diagnosed, the next step is to forecast its progression and potential impact. Prognosis typically relies on:
\begin{itemize}
    \item Expert Knowledge: Veterinarians draw on empirical observations and scientific experimentations (detail more how they build knowledge pieces by pieces) knowledge, to predict the disease trajectory.
    \item Historical Data: Farmer can also rely on past events and past decision outcomes to help inform the likely spread of the disease. 
    % though variability between farms or over time can challenge generalization and accuracy.
\end{itemize}

\paragraph{Recommendation of Control Measures.} Based on the diagnostic and prognostic assessments, appropriate control measures can be devised. These measures are intended to:
\begin{itemize}
    \item Reduce the spread of the disease within and between farms.
    \item Minimize production losses and safeguard food safety.
    \item Optimize the use of resources such as veterinary expertise, time, and financial investment.
    \item Balance short-term emergency responses with long-term sustainable practices
\end{itemize}

While these conventional methods have proven useful, they come with inherent limitations.

\textit{\textbf{Keywords:}} Diagnosis, clinical appraisal, biological examinations, prognosis, expert knowledge, empirical observations, scientific experimentations, recommendations, control measures.


\subsubsection{Challenges to tackle}

\paragraph{Limitations in Diagnosis:} 
\begin{itemize}
    \item Subjectivity and Inconsistency: Visual and manual examinations by farmers or experts, though common, are inherently subjective. Different observers may interpret clinical signs in various ways, leading to inconsistent diagnoses.
    \item Scalability Issues: Manual diagnosis becomes increasingly impractical as farm sizes grow or when frequent assessments are necessary. This limitation is critical as rising demand requires more rapid and consistent monitoring.
    \item Intrusiveness and Delay in Biological Testing: Biological examinations (e.g., blood sampling and PCR testing) offer greater precision but are invasive and time-consuming. The time lag in obtaining results can allow the disease to progress before control measures are implemented.
\end{itemize}

Challenges in Prognosis:
\begin{itemize}
    \item Limited Expertise and Resource Constraints: Relying on expert judgment for prognosis is not scalable. The shortage of veterinarians, especially in rural areas, exacerbates the challenge of providing timely and accurate prognoses across large populations.
\end{itemize}

Addressing these challenges is essential for enhancing the reliability and scalability of disease management strategies. 

\textit{\textbf{Keywords:}} subjectivity, scalability, shortage of expertise, costly, laborious, time-consuming, animal stress, repetitive

% il faudrait que je trouve un endroit où mettre les défifintiosn de certains mots: metapopulation par exemple ?  mais c'est mieux de l'insérer dans le texte si c'est facilemetn possible. 
% je peux jeter un coup d'oel sur l'article de pauline pour avoir d'autres argument
% et regrader de le docs de maud si on parle de méthode traditionnele de control des maladies infectieuse: éleveur comme vétos.
% je vais devoir chercher les chiffres pour argumenter certains faits. example: ethical concerns...

% my old notes
% - il faut bien préciser que c'est ce qui est le plus répandu car c'est le minimum et c'est à la portée de tous mais elle a également ces limites.
% - observation des eleveurs et des experts
% - estimation des états de santé ... 
% - doses des traitement non adaptés: phénomène de resistance des bactéries
% - la santé animale qui prend chère
% - méthode purement empiique qui mais qui peut couter chère
% - d'un environnement à l'autre ce n'est pas toujours les mêmes stratégies qui peuvent être appliquées
% - scalable ? update of the knowledge ? transferability to other issues ? 
% - manual job that is not consistant but it is the baseline that has always been done
% - le problème c'est que non seuelemnt les eleveurs seuls ne possèdent pas toute la connaissance véto et/ou de modélisateurs pour optimiser de manière rigoureuse et stable les .... (lire l'article de pualine pour prouver pourqoui cees méthodes ne peuvent pas faire long terme tel quels s'ils ne sont pas automatisé. (ils ne sont pas mauvais, mais iil faut une pratique standardisé)

% - conclure sur la self-suffisiancy ou susbtainability des filières par rapport aux maladies respiratoires. 
% - Tacler tous ces problèmes nécessites la mise en place de méthodes qui sont repoductible, compréhensible/interpretable, facilement utilisable, et qui sont construit en impliquant la connaissances des tous les acteurs de la filière concernées. 

% - L'idéée est d'avoir des méthodes qui peuvent aider à aléger la tâche aux éleveurs afin qu'ils consacrent leur energies, argent, temps et efforts sur des problèmes à plus forte valeur ajoutés.  
% peuvent nous aider à mimer et faire mieux ou de manière fiable ce ce qu'on.

%----------------------------------------------------------------------------------------
%	SECTION 
%----------------------------------------------------------------------------------------
\clearpage 


\section{Artificial intelligence}

A note on history (un rappel de l'ia, les grands courants... ce qu'on voit tous les jours ne reflètent qu'une partie de l'iceberg imergé ??? bref

ou est-ce que nous on se situe par rapport aux DL, modèle epi et modèle hybride.
Ensuite je me suis intéressé à tel et tel partie (par rapoport aux choses faites dans les modèles hybrides)


Rajouter stéto dans les perspectives (discussions)


\subsection{Precision agriculture: hardware and sensors}

Precision agriculture represents an innovative approach to farming that focuses on optimizing inputs (such as feed, medication, and energy) while maximizing outputs, including net profit and environmental sustainability. This paradigm is rapidly being adopted in livestock farming to enhance decision-making and operational efficiency. At its core, precision agriculture is supported by a robust infrastructure consisting of both hardware and software components. In this subsection, we concentrate on the sensors that underpin data collection:


\begin{itemize}
    \item Behavioural Data: For instance, accelerometers can record movement patterns of individual animals.
    \item Positional Data: Electronic identification systems (such as Eartags) track animal locations and movement across the farm.
    \item Biological Data: Devices like intraruminal boluses can continuously monitor physiological metrics such as body temperature.
    \item Environmental Data: Connected sensors measure ambient conditions, such as temperature and humidity, which can influence animal health.
    \item Operational Data: Information regarding feeding schedules and management practices can also be recorded to provide a comprehensive picture of the farming environment.
\end{itemize}

One of the major benefits of these sensors is their configurability. They can be tailored to collect data at various spatial scales—from detailed individual observations to broader herd-level trends—and temporal resolutions, whether through continuous real-time monitoring or periodic measurements. This scalability ensures that both micro-level behaviours and macro-level patterns are captured effectively. 

By leveraging these advanced sensor systems, precision agriculture can provide a more objective, scalable, and comprehensive means of monitoring livestock health, thereby addressing some of the inherent limitations of conventional observational methods.

Although sensors deliver raw observational data that is closely linked to the emergence and spread of infectious diseases, a conceptual gap remains. The challenge lies in processing this unstructured, multi-dimensional data to extract meaningful insights and translate them into actionable recommendations.

\textit{\textbf{Keywords:}} sensors, automated data collection, multi-modal datasets, scalable, customizable.

% if you have more examples of sensors used in precision agriculture, then i shoudl insert it here. I don't want to mention anyy modelling strategies here nor do i want to mention any, just want to stay limited on the sensors and hardware so do you have any more ideas ? 

\subsection{Epidemiological modelling: Stochastic mechanistic models}

Epidemiological models serve as valuable tools for understanding, predicting, and controlling the spread of infectious diseases. By combining theoretical knowledge of disease transmission dynamics with empirical observations, these models provide insights that guide public health and veterinary interventions.

\paragraph{Fundamental Concepts and Historical Context.} Epidemiological modelling has deep roots in classical infectious disease research, with early examples focusing on simple compartmental models (e.g., SIR, SEIR). Over time, these models have evolved to capture more nuanced disease dynamics and real-world complexities. They explicitly represent the biological and ecological processes governing infection spread, such as contact rates, recovery patterns, and immunity mechanisms. By embedding domain knowledge, mechanistic models are interpretable and allow researchers to dissect how various factors influence disease trajectories.

\paragraph{Stochastic Mechanistic Models.} Stochastic approaches extend mechanistic models by incorporating random events—such as unpredictable contact opportunities, mutations, or interventions—thereby accounting for variability in disease spread. Key features include granular representation of uncertainty (Randomness in infection times and transmission events is captured, leading to a range of possible outcomes rather than a single deterministic trajectory). Stochastic models can better handle heterogeneous populations (e.g., different farms, varying animal susceptibility), especially when data are limited or events are sporadic. Scenario Exploration : Researchers can run multiple simulations under different assumptions or interventions, providing valuable insights into potential disease outcomes without requiring large-scale, real-world experimentation.

Interpretability and Biological Realism: By embedding expert knowledge and explicit mechanisms, these models help highlight the interactions and feedback loops that drive disease progression.

Hypothesis Testing and Policy Guidance: Stochastic mechanistic models allow for the evaluation of “what-if” scenarios, enabling more informed decisions on control measures—such as vaccination strategies or biosecurity protocols.

Although powerful, stochastic mechanistic models can become mathematically complex.

\textit{\textbf{Keywords:}} epidemiological stochastic mechanistic models, explicit representation of underlying mechanisms, incorporation of as much knowledge as we want, explainable predictions and insights, evidence-based recommendations, extrapolation


\subsubsection{Identifiability and observability}

Identifiability and observability are fundamental concepts in epidemiological modelling and statistical inference. They describe our ability to determine meaningful information about a system based solely on available observations (relire l'article de nik cunniffe et frederic hamelin)

Identifiability: Refers to whether the parameters of a model can be uniquely and accurately estimated from the available observations. In epidemiological modelling, a model is identifiable if different parameter values always lead to distinct observational outcomes. A lack of identifiability implies multiple plausible explanations exist for the same observed data, making reliable inference difficult.

Observability: Closely related to identifiability, observability refers specifically to the possibility of inferring the internal state of a system from available observations. An observable epidemiological model allows researchers to infer hidden states (e.g., the number of infected or susceptible animals at a given time) from external observations (e.g., clinical signs, ultrasound imaging, sensor data).

The challenges of identifiability and observability are central when modelling livestock diseases, as they directly affect the reliability of predictions, diagnoses, and recommended interventions. When parameters in a mechanistic epidemiological model cannot be reliably identified, the predictions become uncertain, undermining effective disease management.

Challenges of using epidemiological mechanistic models in Livestock farming:

\begin{itemize}
    \item Parameter inference, conversely, aims to estimate model parameters directly from context-specific observational data. It typically provides more precise results tailored to the studied population or outbreak but requires sufficient data quantity and quality, as well as more computational resources and advanced inferential methods. Real-world epidemiological models often have many parameters. As model complexity increases, the parameter space grows exponentially, a phenomenon known as the "curse of dimensionality." This makes it increasingly difficult to accurately infer parameters from observational data, especially when the quantity or quality of data is limited. 
    \item Manual Processing and Scalability Issues: often, methods including visual analysis or expert-driven parameter selection, are time-consuming and hard to scale. Manual identification of observable patterns or features from raw data becomes impractical when dealing with extensive datasets, particularly data derived from videos, audio recordings, or continuous sensor measurements. Livestock observational data, such as ultrasound imaging or audio recordings of animal coughs, often contain high levels of noise and unstructured information. Such data are challenging to interpretate even by experts, severely limiting scalability and accuracy.
\end{itemize}

\textit{\textbf{Keywords:}} Parameter identification, fitting to non-structured observations, differentiating models, manual model fitting does not scale-up.

Even thought these modelling approach are arguably the best to study and control complex diseases, they still have a few but critical challenges to be tackled. Being able to exploit their knowledge on non-structure observations (which are very frequent in the real-world) would also make them more widely used. 


\subsection{Machine learning modelling: Deep learning models}
Machine learning is a subfield of artificial intelligence involving algorithms that allow computers to learn patterns or relationships directly from data. The defining characteristic of machine learning models is their ability to improve their predictive performance through exposure to increasing amounts of data, rather than through explicitly programmed instructions. Within machine learning, deep learning models represent a sophisticated subclass, specifically designed to handle complex, high-dimensional, and unstructured data. Deep learning employs artificial neural networks—computational models inspired by biological neural systems—which consist of interconnected processing units called neurons arranged in layered architectures. This hierarchical structure enables deep learning models to learn intricate data representations at multiple abstraction levels.

In this thesis, we give extra attention to deep learning architectures, for their high performance in data-mining, pattern extraction in unstructured data which are an efficient format for storing high dimensions observations in their closest form to the original state: "one image contains a thousand words". 
Deep learning models are exceptionally powerful in tasks requiring automatic pattern extraction, and they are frequently described as universal approximators, capable of capturing non-linear relationships and subtle features from large and varied datasets. Typical examples include:

\begin{itemize}
    \item Image classification: Identifying objects or diseases from medical or veterinary images (e.g., ultrasound images).
    \item Natural language processing (NLP): Understanding and generating human language, such as sentiment analysis and text classification.
    \item Speech recognition: Automatically transcribing spoken audio signals.
\end{itemize}

There are methods implemented to enhance the explainability of deep learning models (XAI):
\begin{itemize}
    \item Grad-Cam and LIME produce saliency maps. They provide visual explanations by highlighting important regions of an input image that contribute to a neural network's prediction. Grad-CAM is specific to convolutional neural networks (CNNs) and generates heatmaps by computing gradients of the predicted class score relative to the CNN's intermediate convolutional layers, emphasizing regions that strongly influence the prediction. In contrast, LIME is model-agnostic, working by creating local perturbations around the input data, fitting a simple interpretable model (such as a linear regression) to approximate the behaviour of the original model, and highlighting input regions that most significantly affect the prediction. They are several other methods, such as DeepLIFT (Deep Learning Important Features, SmoothGrad, occlusion sensitivity). 
    \item Uncertainty quantification such as Bayesian deep learning or conformal predictions. There is no rivalry. They are radically different frameworks achieving very different things. CP is about calibrating predictions, while the Bayesian framework is about quantifying uncertainty of various things from the model, parameters, and consequently, the predictions. The typical guarantee you get from the Bayesian framework is that your estimates/predictions achieve the smallest average loss over the prior. That means Bayesian estimators are optimal estimators in terms of average performance, meaning they will be accurate, but do not guarantee coverage. CP does nothing about the accuracy of your estimator but gives you coverage. You can combine both and have the best of both worlds. One can understand this from a PAC-Bayesian perspective: CP is related to the empirical risk, while BDL is related to the KL divergence between the posterior and prior. 
\end{itemize}

These examples illustrate the flexibility and adaptability of deep learning across multiple application domains, including precision agriculture and epidemiological diagnostics. Despite their remarkable capabilities, deep learning models present certain limitations critical in epidemiological and livestock farming contexts. 

\textit{\textbf{Keywords:}} Deep learning, automated pattern extraction, handles unstructured observation, most powerful function approximator, intrapolation, uncertainty quantification, XAI


\subsubsection{Limitations and Challenges of classical Deep Learning Models}

% Interpolative vs extrapolative knowledge

Deep learning methods typically assume the data to be independently and identically distributed (iid), meaning each data point is generated from the same probability distribution and independently from one another. Deviations from this assumption can significantly degrade model performance and generalization capabilities.

% \begin{itemize}
%     \item Lack of Interpretability and Biophysical Meaning: Deep learning models are often described as "black-box" models due to their complexity and limited interpretability. The learned parameters (weights and biases) generally do not possess direct biological or mechanistic meaning, making it difficult to understand why a model produces a certain prediction.
%     \item Overfitting and Generalization: With high capacity and flexibility comes the risk of overfitting, where the model learns dataset-specific noise or irrelevant patterns rather than meaningful generalizable trends. Consequently, achieving robust generalization performance beyond the training set can be challenging.
%     \item Requirement for Extensive and Representative Data: Deep learning models rely heavily on datasets that comprehensively represent the underlying phenomena of interest. Insufficient or biased training data can lead to suboptimal generalization, especially in disease diagnostics, where precise and representative datasets are rare, difficult to collect, or ethically challenging to generate (e.g., intentionally inducing disease outbreaks).
%     \item Sensitivity to Distribution Shifts (Out-of-Distribution Data): Deep learning models typically struggle when encountering data outside their training distribution (Out-of-Distribution, OOD). This limitation is particularly pertinent in epidemiology, where the appearance of previously unobserved disease variants or changes in environmental conditions frequently occur.
%     \item these methods can also quickly become very resource intensive particularly when dealing with complex task as often the architecture also get more complex, requiring bigger infrastructure (gpu, cpu) for training or inference, these also has an impact on environmental management.
% \end{itemize}

XAI methods are a way to measure and understand why predictions are made, They are usually are employed for outliers detection, data drift, ...  however they cannot be used to precisely forecast extreme scenarios. 

Each of these methodologies individually addresses certain specific aspects of infectious disease modelling effectively, but they also carry inherent limitations when applied alone:
\begin{itemize}
    \item Deep learning models excel at extracting intricate patterns from large-scale or high-dimensional observational data (e.g., sensor-derived data such as audio, video, or biological signals). Their data-driven nature allows them to effectively capture context-specific and complex relationships without explicit theoretical assumptions. Nevertheless, their main limitation lies in their lack of explicit biological or theoretical interpretability, limiting their capability to reliably extrapolate to unseen scenarios or out-of-distribution contexts.
    \item Stochastic mechanistic epidemiological models explicitly encode theoretical or biological mechanisms (infection transmission, recovery rates, or disease progression) using interpretable parameters with clearly defined biological meanings. These models can thus generate hypotheses, facilitate understanding of causal dynamics, and provide robust extrapolation to scenarios not explicitly observed in the data. However, their main challenge lies in effectively incorporating detailed, unstructured, and context-specific observational data (such as sensor data), limiting their practical utility and responsiveness to specific epidemiological contexts.
\end{itemize}

% Even thought these model are widely used to process complex datasets, they are hardly used to explore novel situations or predict task that are inherently has multiple complex relationships (so hard to gather a representative dataset). these challenges essentially stem from the lack of explicit knowledge incorporation (example hallucinations in LLMs). They are in their need of feature that could  enhance interpretability, improve generalization in OOD settings, and facilitate more accurate and meaningful insights by explicitly incorporating domain-specific knowledge into the learning process.


% do models like PINN require more data ? 
% otherwise the advantage of what we are proposing will be on the structure i guess. A modular method, that could require a bit less since we are training a a less complex task. at yet maybe as performant ? 

% prmireè formulation: a methodology that can be used to process non-structure datasets and still easily be used to extrapolate towards unknown complex scenarios.
% second thing incorporate more knowledge into the observation that are gathered because sometimes the observations are noisy and don't represent the global mechanisms. 

\textit{\textbf{Keywords:}} Generalization, forecast extreme scenarios, input dataset has to be iid, lack of biophysical meaning of parameters, black-box modelling, data intensive for complex tasks, resource intensive (GPU, electricity, materials...) for complex tasks, false positive, false negative, extrapolation


\subsection{Hybrid modelling: machine learning and epidemiological models}

[Raccourci cette partie pour garder le tableau, le plus important c'est de montrer les objectifs (summarize), les types de couplages (summarize), montrer les axes que nous mettons en avant]
En terme d'objectif voici les cases que nous cochons: model parameterizination, disease intervention assessemnt and optimization, restrospective epidemic course analysis, infectious disease forecasting (par contre préciser qu'ici en réalité nous avons n'avons pas fait de forecast dans le sens statistique du terme jusqu'à aller à l'évaluation des performances, transmission inference, outbreak detection]

[Résumé les objectifs en un paragraphe. Etendre par contre le type de d'intégration et comment c'est fait. mettre égalemenet le rôle que joue chaque partie ?] [the roles are long-term forecast (or prognose) or short-term predictions]

[numerotez les approches d'intégration, ça va m'être utile pour la suite]

This section serves as a concise narrative review of various methodologies for integrating machine learning models with epidemiological models, offering a critical perspective on their limitations. Addressing these limitations is central to the contributions of this thesis.

This section presumes that the reader possesses a foundational understanding of ensemble techniques in modelling, specifically bagging, boosting, and stacking. They are pivotal in enhancing the global performance by combining multiple models. These methods aim to reduce errors, improve accuracy, and increase the robustness of predictions. Stacking Combines outputs from several distinct models using an additional "meta-model," learning optimal combinations for improved predictions. Boosting refers to sequentially training multiple weak predictive models, each aiming to correct errors from previous models, thus progressively improving overall prediction accuracy. Bagging (or Bootstrap Aggregating) refers to parallel training of multiple unique models (while stacking used distinct models, bagging uses multiple instances but of the same model) on random subsets of data (with replacement), thereby reducing prediction variance and mitigating over-fitting.

In the scientific literature, hybrid modelling approaches integrating deep learning and mechanistic models have gained attention, especially during epidemiological crises such as the COVID-19 pandemic. Several prominent methodologies in this domain include:

En sciences, toute tentative de tirer des conclusions généralisables nécessite un
modèle, i.e. une version abstraite de la réalité \cite{McCallum2008}. Cette représentation
abstraite est dotée d’un objectif (a model is a "purposeful representation", 
\cite{Starfield1990}, qui requiert un certain degré de simplification.

The purposes of the hybrid models currently found in the litterature can be grouped into 6 mutually exclusive groups, indicating that a single integrated model serve multiple application areas \cite{Ye2025} :
\begin{itemize}
    \item Infectious disease forecasting: Predicting the future spread or trajectory of infectious diseases by combining AI's data-mining strengths with the explanatory power of mechanistic epidemiological models. \cite{Poirier2020, Fan2022, Garner2016, Kharazmi2021, Barmparis2022}
    \item Model Parameterization and Calibration: Determining and refining model parameters (e.g., disease transmissibility, contact rates) accurately and efficiently, often by extracting additional insights from diverse and complex datasets. \cite{Tuarob2015, SolaresHernandez2023, Reiker2021, Jorgensen2022, Raissi2019}
    \item Disease Intervention Assessment and Optimization: Evaluating and identifying optimal strategies for interventions (such as vaccinations, social distancing, or lockdowns) to minimize disease spread and impact, utilizing methods like reinforcement learning and game theory. \cite{Thiagarajan2022, Bertsimas2021, Aurell2022, Janko2023}
    \item Retrospective Epidemic Course Analysis: Understanding past epidemics by analyzing historical data, identifying critical factors influencing transmission dynamics, and informing future outbreak preparedness. \cite{Ruth2022, Santermans2015, Sanson2022, Weyant2023}
    \item Transmission Inference: Inferring hidden patterns, source origins, and transmission networks using AI techniques trained on data from mechanistic epidemiological simulations or real-world data. \cite{Ling2022, Colijn2014, Vilar2023}
    \item Outbreak Detection:Quickly identifying new outbreaks or unusual surges in cases, enabling timely public health responses through AI-enhanced analysis of varied data sources, such as medical reports or social media content.
\end{itemize}


Methodologicaly-wise, these models can also grouped into nine primary integration approach:
\begin{itemize}
    \item AI-augmented epidemiological models: These models directly replace (or augment) specific components (e.g., parameters or even entire sub-models) of the mechanistic framework with machine learning outputs. In many cases, the machine learning component is trained end‐to‐end alongside the mechanistic model or is integrated into the numerical solver (réfs heres). In an end-to-end configuration, the machine learning module’s weights are updated simultaneously with the mechanistic model’s parameters, creating a tightly coupled structure that learns both representation and dynamics jointly. 
    \item Epidemiological models with improved observational data: In these approaches, AI is applied to augment or “clean” raw data inputs before they feed into the mechanistic model. Techniques include using machine learning to extract features from social media, search trends, or satellite imagery(réfs heres). Machine learning operates as a pre-processor and the mechanistic model then uses the enhanced, higher-quality data for calibration and simulation. This is typically a weakly coupled integration because the machine learning module’s output is treated as a fixed input to the downstream mechanistic model rather than being jointly optimized with it.
    \item Physics-infromed neural networks (PINNs) and Epidemiology-aware AI models (EAAMs): These models embed the differential equations governing disease transmission directly into the loss function of the neural network (réfs heres). PINNs are effective when generalised to some PDEs, in particular nonlocal or high-dimensional PDEs, for which traditional solvers are computationally expensive. However, \cite{Kidger2021}  emphasises that this is a distinct notion to AI-augmented epidemiological models (AIEM). AIEMs use neural networks to specify differential equations. PINNs uses neural networks to obtain solutions to prespecified differential equations. This distinction is a common point of confusion, especially as the PDE equivalent of PINNs is sometimes referred to as a `neural partial differential equation'.
    \item Machine learning models incorporating epidemiological input features: In this configuration, traditional machine learning architectures (like recurrent neural networks or graph neural networks) are augmented by explicitly incorporating epidemiological features (e.g., contact rates, incubation periods) as additional inputs or via modified architectures. This represents a moderate or weak coupling since the epidemiological knowledge is “injected” as additional features rather than being deeply embedded in the model’s structure or loss function.
    \item Surrogate modelling/synthetically-trained ai models: Surrogate models use machine learning to approximate the outputs of computationally expensive mechanistic models. Typically, synthetic datasets are generated from simulations of the mechanistic model, and an machine learning model (often a neural network) is then trained on these data. The coupling here is often “loose” or weak because the AI surrogate is a stand-alone approximation that is later used in place of the full mechanistic model.
    \item Ensemble Learning Frameworks: Ensemble methods combine predictions from mechanistic models and machine learning models (or multiple ML models) using techniques like weighted averaging, stacking. The coupling is relatively loose since each model is usually developed independently and then combined at a later stage.
    \item Bayesian Neural Networks: These hybrid approaches integrate Bayesian inference with deep learning to quantify uncertainty in parameter estimation. In the epidemiological context, Bayesian neural networks are used to learn parameters while simultaneously providing credible intervals for model outputs.The coupling is inherent in the training process as the posterior distribution is approximated jointly using both data likelihood and prior mechanistic information.
    \item ML-Enhanced Optimization Frameworks: These methods use AI algorithms—such as reinforcement learning, optimal control, or Markov decision processes—to optimize intervention strategies based on outputs from mechanistic models.In an end-to-end framework, the AI agent interacts with the mechanistic simulation in a closed loop, receiving feedback and updating its strategy iteratively (tight coupling).Alternatively, the optimization can be performed in a decoupled fashion where the mechanistic model is used to generate scenarios and the AI module optimizes the intervention based on pre-computed results (weaker coupling).
    \item Cluster-Based Transmission Analysis Frameworks: This approach uses unsupervised learning methods (such as k-means clustering) to segment populations or regions with similar transmission characteristics. Once clusters are identified, mechanistic models are applied to each subgroup to refine forecasts or intervention strategies.
\end{itemize}

These findings can be summarized into a table \ref{tab:integration_methods}:

\begin{table}[htbp]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|p{4cm}|p{5cm}|p{3cm}|p{4cm}|p{4cm}|p{4cm}|p{6cm}|}
\hline
\textbf{Integration approach} & \textbf{Objective} & \textbf{Coupling Type} & \textbf{ML Architecture} & \textbf{Epidemiological Architecture} & \textbf{Aggregation type} & \textbf{Full Reference} \\ \hline

AI-augmented epidemiological models &
Infectious disease forecasting, Model parametrisation and calibration &
Strong coupling &
Neural networks, regression, tree-based models replacing parts of the traditional models for parameter estimation &
Mechanistic epidemiological models with components replaced or augmented with by ML &
End-to-end training &
Arik et al. (2021) \\ \hline

Epidemiological models with improved observational data &
Outbreak detection, model parametrization and calibration &
Weak coupling &
methods like SVM , tree-based, NLP, MLP algorithms used to extract and refine features from non tradictional surveillance data (text) &
Bayesian epidemiological frameworks (or classical compartmental or mechanistic models updated using enhanced observational inputs derived from diverse data sources) &
Stacking, Bayesian model selection &
Tuarob et al. (2015), Rosato et al.(2023), Kandula et al.(2019) \\ \hline

PINNs (physics-informed neural networks) / EAAMs (Epidemiology-Aware AI models) &
Model parametrisation and calibration, Infectious Disease Forecasting &
Strong coupling &
Neural networks (DNN, CNN, RNN, GNN) that incorporate physics-based loss functions (combining data loss and residual loss enforcing differential equations) &
Differential equations-based mechanistic models(Compartmental, renewal equations e.g: SIR, etc) &
End-to-End training &
Cai, M., Karniadakis, G. E., \& Li, C. (2022). Fractional SEIR model and data-driven predictions of COVID-19 dynamics of Omicron variant. \textit{Chaos}, 32, 071101; 

Ghosh, S. et al. (2023). Understanding the implications of under-reporting, vaccine efficiency, and social behaviour on the post-pandemic spread using physics informed neural networks. \textit{PLoS One}, 18, e0290368. Barmparis \& Tsironis (2022), De rosa et al. (2023) \\ \hline

AI models incorporating epidemiological input features &
Infectious disease forecasting &
Weak coupling &
ML models (e.g: FFN, RNN, Tree-based models) that integrate epidemioloigcal indicators as part of their input features &
Mechanistic epidemiological models supplying key epidemiological metrics (e.g contact rates, parameters) as inputs to the AI model &
Feature-level integration &
- \\ \hline

Surrogate modelling / synthetically-trained AI models &
Model parametrisation and calibration, Disease intervention assessment and optimization, Retrospective epidemic course analysis &
Weak coupling &
Lightweight neural networks or 
Tree-based models trained on synthetic datasets &
simulation-based mechanistic models (compartmental or Agent-based models) for synthetic generation &
Surrogate modelling &
Anirudh, R. et al. (2022). Accurate Calibration of Agent-based Epidemiological Models with Neural Network Surrogates. \textit{Proceedings of the 1st Workshop on Healthcare AI and COVID-19, ICML 2022}, vol. 184, pp. 54–62, PMLR. \\ \hline

Ensemble learning frameworks &
Infectious disease forecasting &
Weak coupling &
combinations of models such as LTSM, stacking and boosting methods to merge prediction from different sources &
Traditional mechanistic models whose forecasts are combined with AI-based predictions &
Weighted averaging, stacking, boosting &
Kandula et al. (2018), Adiga et al. (2021), Nadler et al. (2020),
Maniamfu \& Kameyama (2023), Delli compagni et al. (2022)\\ \hline

Bayesian neural networks &
Model parameterisation and calibration &
Strong coupling &
Bayesian neural networks enabling uncertainty quantification in parameters estimation &
mechanistic model providing a probabilistic structure (e.g SIR models with uncertainty in parameters) Agent-based models, Ordinary differential equations &
Bayesian inference &
Kwok, W. M., Streftaris, G., \& Dass, S. C. (2023). Laplace based Bayesian inference for ordinary differential equation models using regularized artificial neural networks. \textit{Stat. Comput.}, 33, 124. \\ \hline

AI-enhanced optimization frameworks &
Disease intervention assessment and optimization &
Strong coupling &
Reinforcement learning, neural networks, optimal control neural networks, and game-theoretic models designed to learn optimal intervention strategies &
often agent-based epidemiological models that serve as environments for optimization and decision-making &
Reinforcement learning, optimal control &
Yien et al. (2023), Jian et al. (2023), Malloy \& Brandeau (2022), Zang et al. (2022), Janko et al. (2023) \\ \hline

Cluster-based transmission analysis frameworks &
Infectious disease forecasting, Transmission inference &
Weak coupling &
Clustering algorithm (k-means, Hierarchical clustering...) and network analysis methods (Graph neural networks) to identify and analyze clusters in transmission data &
Network-based or agent-based models that capture transmission dynamics and suppot the identification of key transmission clusters
Individual-based models &
Clustering & - \\ \hline
\end{tabular}%
}
\caption{Summary of nine primary integration method types.}
\label{tab:integration_methods}
\end{table}

% je pourrais rajouter deux colonnes pour les avantages et les limites ou les architectures utilisés pour le deep comme pour les modèles epidelio

\begin{table}[htbp]
\centering
\resizebox{\textwidth}{!}{%
\begin{tabular}{|p{4cm}|p{5cm}|p{3cm}|p{4cm}|p{4cm}|p{4cm}|p{6cm}|}
\hline
\textbf{Integration approach} & \textbf{Objective} & \textbf{Description} & \textbf{training process}  & \textbf{References} \\ \hline


PINNs (Physics-Informed Neural Networks) &
Model parametrisation and calibration, infectious disease forecasting &
A neural network is trained to approximate the state variables (such as susceptible, infected, and recovered populations) as functions of time. The key design feature is the incorporation of the governing differential equations (for example, those from a classical SIR model) directly into the loss function. This composite loss includes a data loss term (comparing network predictions to observed time-series data) and a residual loss term that forces the solution to satisfy the epidemiological equations &

Training Process: Gradient-based optimization (using methods like Adam or SGD) is applied to minimize the combined loss function, ensuring that both the data fit and the differential constraints are respected &

Original Citations: References [32, 59–66] \\ \hline

PINNs (Physics-Informed Neural Networks) &
Model parametrisation and calibration, infectious disease forecasting &
A neural network is trained to approximate the state variables (such as susceptible, infected, and recovered populations) as functions of time. The key design feature is the incorporation of the governing differential equations (for example, those from a classical SIR model) directly into the loss function. This composite loss includes a data loss term (comparing network predictions to observed time-series data) and a residual loss term that forces the solution to satisfy the epidemiological equations &

Training Process: Gradient-based optimization (using methods like Adam or SGD) is applied to minimize the combined loss function, ensuring that both the data fit and the differential constraints are respected &

Original Citations: References [32, 59–66] \\ \hline

\end{tabular}%
}
\caption{Summary of nine primary integration method types.}
\label{tab:integration_methods}
\end{table}







\begin{itemize}
    \item Physics-Informed Neural Networks (PINNs): PINNs integrate known governing equations (e.g., differential equations describing infectious dynamics) as soft constraints into neural networks during training. The neural network is guided by both observed data and theoretical knowledge, ensuring physically plausible predictions and improved extrapolation beyond the training distribution.
    \item Neural Differential Equations (Neural ODEs/SDEs): Neural ODEs or SDEs embed neural networks into differential equations, allowing the modelling of dynamic processes through continuous-time representations. Such integration directly captures temporal dynamics, offering more flexibility and interpretability compared to classical discrete-time deep learning methods.
\end{itemize}

Despite their promising advantages, hybrid and ensemble models have inherent challenges:
% Despite notable advancement in the scientific community, there is a conceptual gap, between the raw unstructured livestock data that can be captured and the strong theoretical knowledge we have of the complex underlying infectious mechanisms. This basically allows to ground the theoretical knowledge we have on a complex mechanisms to the context-specific observations, hence harness the best of both world. 

\textit{\textbf{Keywords:}} Ensemble learning, Physics-informed neural networks, Neural differential equations


Each ensemble strategy presents specific strengths and limitations, related to computational complexity, interpretability, and predictive power. 
research gaps:
\begin{itemize}
    \item Generalizability: The majority of integrated models focus on diseases like COVID-19, limiting applicability to other infectious diseases with indirect or complex transmission routes.
    \item Expansion beyond COVID-19 to other diseases with indirect transmission mechanisms.
    \item Data Limitations: Many studies depend on synthetic or limited real-world data, highlighting the need for richer, real-time data integration (e.g., satellite imagery, social media data). - we have gathered in this thesis a real-wolrd multi-modal dataset (...)
\end{itemize}

\subsubsection{Maintainability}

    \begin{itemize}
        \item strongly coupled methods requires the builder to be deeply understand the problematic of domain in order to implement the models. (full stack vs frontend or backend). So retraining for example would require the whole system to be trained and in the long-term, it would also require sort of full-stack builders. 
        
        \item strongly coupled methods would still require a  certain quantity of data since in order for example for the deep learning model to fine-tune the mechanistic model (neural differential equations). Usually ensembled model (so more parameters) would require more dimensions in order to fit all the parameters (curse of dimensionality)
    \end{itemize}

% le domaine est très large, mais nous on va se focaliser uniquement sur le couplage deep et modèle méchaniste de façon générale. 

% sachant que nous c'est une forme de boosting, sans trop de redondances,) et surtout découplé (faiblement couplé) - ceci sont des couplages forts mais comme un site web à egalement ces limites. comme un développeur full stack - parler rapidement des NDE, des pinns qui sont juste une sous familles des NDE.

\textit{\textbf{Keywords:}} ...



%----------------------------------------------------------------------------------------
%	SECTION 
%----------------------------------------------------------------------------------------
\clearpage 

\input{chapters/thesis_outline}
